---
title: "Lab 08 - Text Mining/NLP"
output: github_document
always_allow_html: true
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(eval = T, include  = T)
```

# Learning goals

- Use `unnest_tokens()` and `unnest_ngrams()` to extract tokens and ngrams from text
- Use dplyr and ggplot2 to analyze and visualize text data
- Try a theme model using `topicmodels`

# Lab description

For this lab we will be working with the medical record transcriptions from https://www.mtsamples.com/ available at https://github.com/JSC370/JSC370-2025/tree/main/data/medical_transcriptions.

# Deliverables

1. Questions 1-7 answered, knit to pdf or html output uploaded to Quercus.

2. Render the Rmarkdown document using `github_document` and add it to your github site. Add link to github site in your html.
  1. [Github](https://github.com/e1444/JSC370-labs)


### Setup packages

You should load in `tidyverse`, (or `data.table`), `tidytext`, `wordcloud2`, `tm`, and `topicmodels`.


## Read in the Medical Transcriptions

Loading in reference transcription samples from https://www.mtsamples.com/

```{r message = FALSE}
library(tidytext)
library(tidyverse)
library(wordcloud2)
library(tm)
library(topicmodels)

mt_samples <- read_csv("https://raw.githubusercontent.com/JSC370/JSC370-2025/main/data/medical_transcriptions/mtsamples.csv")
mt_samples <- mt_samples |>
  select(description, medical_specialty, transcription)

head(mt_samples)
```

---

## Question 1: What specialties do we have?

We can use `count()` from `dplyr` to figure out how many different medical specialties are in the data. Are these categories related? overlapping? evenly distributed? Make a bar plot.

```{r}
mt_samples |>
  count(medical_specialty) |>
  ggplot(aes(x = reorder(medical_specialty, n), y = n)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  labs(x = "Specialty", y = "Count", title = "Medical Specialties")
```

The data is not evenly distributed, with surgery being the most representative specialty by a wide margin, making up nearly a third of th data. There are 30 categories in total, most of which are medical, but some miscellaneous categories are also present such as "Office Notes".

---

## Question 2: Tokenize

- Tokenize the the words in the `transcription` column
- Count the number of times each token appears
- Visualize the top 20 most frequent words with a bar plot
- Create a word cloud of the top 20 most frequent words

### Explain what we see from this result. Does it makes sense? What insights (if any) do we get?

```{r}
tokens <- mt_samples |>
  select(transcription) |>
  unnest_tokens(word, transcription) |>
  count(word, sort = TRUE)

tokens |>
  slice_max(n, n = 20) |>
  ggplot(aes(x = reorder(word, n), y = n)) +
  geom_bar(stat = "identity") +
  labs(x = "Word", y = "Count", title = "Top 20 Words")

wordcloud2(tokens |> slice_max(n, n = 20))
```

We do not gain any meaningful insights from the word cloud, as the most frequent words are articles, prepositions, and other common words that do not provide any information about the content of our data.

---

## Question 3: Stopwords

- Redo Question 2 but remove stopwords
- Check `stopwords()` library and `stop_words` in `tidytext`
- Use regex to remove numbers as well
- Try customizing your stopwords list to include 3-4 additional words that do not appear informative

### What do we see when you remove stopwords and then when you filter further? Does it give us a better idea of what the text is about?

```{r}
head(stopwords("english"))
length(stopwords("english"))
head(stop_words)

tokens <- mt_samples |>
  select(transcription) |>
  unnest_tokens(word, transcription) |>
  count(word, sort = TRUE) |>
  filter(!word %in% stopwords("english")) |>
  filter(!word %in% c("left", "right", "using", "noted")) |>
  filter(!str_detect(word, "\\d+"))

tokens |>
  slice_max(n, n = 20) |>
  ggplot(aes(x = reorder(word, n), y = n)) +
  geom_bar(stat = "identity") +
  labs(x = "Word", y = "Count", title = "Top 20 Words")

wordcloud2(tokens |> slice_max(n, n = 20))
```

We see some other non-informative words as well as numbers, which we manually filter out. We are left with patient as the most frequent word, which makes sense in the context of our data.

---



## Question 4: ngrams

Repeat question 2, but this time tokenize into bi-grams. How does the result change if you look at tri-grams? Note we need to remove stopwords a little differently. You don't need to recreate the wordclouds.

```{r}
stopwords2 <- c(stopwords("english"), "left", "right", "using", "noted", as.character(0:9))

sw_start <- paste0("^", paste(stopwords2, collapse=" |^"), "$")
sw_end <- paste0("", paste(stopwords2, collapse="$| "), "$")

tokens_bigram <- mt_samples |>
  select(transcription) |>
  unnest_tokens(ngram, transcription, token = "ngrams", n = 2) |>
  filter(!str_detect(ngram, sw_start)) |>  # Remove stopwords at the beginning
  filter(!str_detect(ngram, sw_end)) |>    # Remove stopwords at the end
  count(ngram, sort = TRUE)

head(tokens_bigram)
```

```{r}

tokens_trigram <- mt_samples |>
  select(transcription) |>
  unnest_tokens(ngram, transcription, token = "ngrams", n = 3) |>
  filter(!str_detect(ngram, sw_start)) |>  # Remove stopwords at the beginning
  filter(!str_detect(ngram, sw_end)) |>    # Remove stopwords at the end
  count(ngram, sort = TRUE)

head(tokens_trigram)
```

We observe more informative ngrams. "year old" is the most frequent bigram, and "prepped and draped" is the most frequent trigram. These ngrams provide more information about the content of our data than the single words.

---

## Question 5: Examining words

Using the results from the bigram, pick a word and count the words that appear before and after it, and create a plot of the top 20.

```{r}
library(stringr)
# e.g. patient, blood, preoperative...
# choose "patient"
tokens_bigram2 <- tokens_bigram |>
  filter(str_detect(ngram, "patient .*|.* patient")) |>
    mutate(word = str_remove(ngram, "patient"),
           word = str_remove_all(word, " "))

tokens_bigram2 |>
  slice_max(n, n = 20) |>
  ggplot(aes(x = reorder(word, n), y = n)) +
  geom_bar(stat = "identity") +
  labs(x = "Word", y = "Count", title = "Top 20 Words")

wordcloud2(tokens_bigram2 |> slice_max(n, n = 20) |> select(word, n))
```

---


## Question 6: Words by Specialties

Which words are most used in each of the specialties? You can use `group_by()` and `top_n()` from `dplyr` to have the calculations be done within each specialty. Remember to remove stopwords. How about the 5 most used words?


```{r}
mt_samples |>
  unnest_tokens(word, transcription) |>
  filter(!word %in% stopwords2) |>
  group_by(medical_specialty) |>
  count(word, sort = TRUE) |>
  top_n(1)
```

Amongst the top 10, 9 of the categories have "patient" as the most frequent word, with the exception being radiology, where "normal" is the most frequent.

```{r}
mt_samples |>
  unnest_tokens(word, transcription) |>
  filter(!word %in% stopwords2) |>
  group_by(medical_specialty) |>
  count(word, sort = TRUE) |>
  top_n(5)
```

In the top 5, we see more variation in the most frequent words.

## Question 7: Topic Models

See if there are any themes in the data by using a topic model (LDA). 

- you first need to create a document term matrix
- then you can try the LDA function in `topicmodels`. Try different k values.
- create a facet plot of the results from the LDA (see code from lecture)


```{r}
run_lda <- function(k) {
  transcripts_dtm <- mt_samples |>
    select(transcription) |>
    unnest_tokens(word, transcription) |>
    filter(!word %in% stopwords2) |>
    DocumentTermMatrix()

  transcripts_dtm <- as.matrix(transcripts_dtm)

  transcripts_lda <- LDA(transcripts_dtm, k = k, control = list(seed = 1234))

  top_terms <- 
    tidy(transcripts_lda, matrix = "beta") |>
    group_by(topic) |>
    slice_max(beta, n = 10) |> 
    ungroup() |>
    arrange(topic, -beta)

  top_terms |>
    mutate(term = reorder_within(term, beta, topic)) %>%
    ggplot(aes(beta, term, fill = factor(topic))) +
    geom_col(show.legend = FALSE) +
    facet_wrap(~ topic, scales = "free") +
    theme_bw()+
    scale_y_reordered()
}
```

```{r}
run_lda(3)
```

```{r}
run_lda(5)
```

```{r}
run_lda(7)
```
